---
title: "Getting started with slxr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with slxr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Why SLX

The Spatial-X (SLX) model has the form

$$
y = X\beta + WX\theta + \varepsilon,
$$

where `W` is a spatial weights matrix and `WX` adds spatially-lagged
versions of selected regressors. Wimpy, Whitten, and Williams (2021)
argue that the SLX specification more faithfully reflects typical
political science theories than the more common SAR model, and is much
easier to estimate and interpret: it is plain OLS on an augmented design
matrix.

`slxr` exists because the mechanics of building `W`, multiplying it by
the right columns of `X`, and reporting direct/indirect/total effects
cleanly is more friction than applied researchers should have to
endure.

## Installation

```{r}
# remotes::install_github("cwimpy/slxr")
library(slxr)
```

## A minimal example

We use the `nc` shapefile shipped with `sf`.

```{r}
library(sf)
nc <- st_read(system.file("shape/nc.shp", package = "sf"))

W <- slx_weights(nc, style = "contiguity")
W
```

Fit an SLX model that spatially lags `BIR74` only:

```{r}
fit <- slx(SID74 ~ BIR74 + NWBIR74,
           data = nc,
           W    = W,
           lag  = "BIR74")

summary(fit)
```

The spatial-lag columns are reported alongside the direct regressors.
To extract clean direct, indirect, and total effects:

```{r}
slx_effects(fit)
```

## Variable-specific weights matrices

The defining feature of Wimpy, Whitten, and Williams (2021) is that
different covariates can spill over through different `W` matrices. For
example, civil war might spill through contiguity while defense-pact
effects spill through an alliance network:

```{r}
W_contig <- slx_weights(countries, style = "contiguity")
W_pact   <- slx_weights(countries, style = "custom",
                         matrix = pact_adjacency)

fit <- slx(defense ~ civil_war + interstate_war + defense_lag,
           data = panel,
           spatial = list(
             civil_war      = W_contig,
             interstate_war = W_contig,
             defense_lag    = list(contig = W_contig, pact = W_pact)
           ))
slx_effects(fit)
```

## Comparison with `lm()` output

Because SLX is OLS on an augmented design matrix, `tidy()` and
`glance()` work as expected:

```{r}
broom::tidy(fit)
broom::glance(fit)

modelsummary::modelsummary(list(OLS = ols_fit, SLX = fit))
```

## References

Wimpy, C., Whitten, G. D., & Williams, L. K. (2021). X Marks the Spot:
Unlocking the Treasure of Spatial-X Models. *Journal of Politics*,
83(2), 722–739.
